---
title: 'Lecture 2: Chapter 3 by Kieran Healy'
author: "Marc Kaufmann"
date: "September 23, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Team Assignments: Choosing First Group

For the first group, I will pick 4 students who will complete a group assignment to share in 2 weeks time (lecture 4). Since it is the first group, I am happy to provide a bit more feedback if needed. So let's do this.

If you want to switch with someone else and someone else is happy to, then please go ahead and swap - just let me know before the weekend.

## Note to the interested student

Try to follow along by typing it yourself, adding comments as you make mistakes or realize things. Write the code out in chunks:

```{r}
```

## How Ggplot Works

```{r, load, warning=FALSE}
library(tidyverse)
```

The code specifies the connections between the variables in the data on one hand and the colors, points, and shapes you see on the screen. These logical connections are called *aesthetic mappings* or simply *aesthetics*.

How to use ggplot:

- `data = gapminder`: Tell it what your data is
- `mapping = aes(...)`: How to map the variables in the data to aesthetics
  - axes, size of points, intensities of colors, which colors, shape of points, lines/points
- Then say what type of plot you want:
  - boxplot, scatterplot, histogram, ...
  - these are called 'geoms' in ggplot's grammar, such as `geom_point()` giving scatter plots
  
  
You link these steps by *literally* adding them together with `+` as we'll see.

**Exercise:** What other types of plots are there? Try to find several more `geom_` functions.  

### Answer:
### The other types of plots include:
### * one variable: 1. f + geom_area(stat="bin"); 2. f + geom_density(kernel="gaussian"); 3. f + geom_histogram(binwidth=5); 4. geom_dotplot(); etc 
### * discrete: b + geom_bar()
### * graphical primitives: 1. geom_polygon; 2. geom_ribbon(); 3. geom_rect
### * two variables: 1. geom_jitter(); 2. geom_quantile(); 3. geom_text(); 4. geom_hex()geom_step().; etc
### * three variables: 1. geom_contour(); 2. geom_tile ()
### the cheatsheet can be found here: [link](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)


## Mappings Link Data to Things You See

```{r}
library(gapminder)
library(ggplot2)
gapminder

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()
```

In detail:

- `data = gapminder` tells ggplot to use gapminder dataset, so if variable names are mentioned, they should be looked up in gapminder
- `mapping = aes(...)` shows that the mapping is a function call. Simply accept that this is how you write it
  - Kieran Healy: "The `mapping = aes(...)` argument *links variables* to *things you will see* on the plot"
- `aes(x = gdpPercap, y = lifeExp)` maps the GDP data onto `x`, which is a known aesthetic (the x-coordinate) and life expectancy data onto `x`
  - `x` and `y` are  predefined names that are used by `ggplot` and friends
  
Importantly, mappings don't say *what* color or shape some variable will have -- rather, it says that a given dataset will be mapped *to* the color or *to* the shape.

```{r, what_is_in_p, eval=FALSE}
str(p)
str(p + geom_point())
```

**Exercise:** Make sure that your knitted version doesn't include all the output from the `str(...)` commands, it's too tedious.  

### Answer
### Added {eval=FALSE}. The knitted version did not contain the output of the str(...)

Finally, we add a *layer*. This says how some data gets turned into concrete visual aspects.

```{r, scatter_plot}
p + geom_point()
p + geom_smooth()
```

**Note:** Both geom's use the same mapping, where the x-axis represents ... and the y-axis ... . But the first one maps the data to individual points, the other one maps it to a smooth line with error ranges.

We get a message that tells us that `geom_smooth()` is using the method = 'gam', so presumably we can use other methods. Let's see if we can figure out which other methods there are.

You may start to see why ggplots way of breaking up tasks is quite powerful: the geometric objects (long for geoms) can all reuse the *same* mapping of data to aesthetics, yet the results are quite different. And if we want later geoms to use different mappings, then we can override them -- but it isn't necessary.

One thing about the data is that most of it is bunched to the left. If we instead used a logarithmic scale, we should be able to spread the data out better.

```{r, scale_coordinates}
p + geom_point() + geom_smooth(method = "lm") + scale_x_log10()
```

**Exercise:** Describe what the `scale_x_log10()` does. Why is it a more evenly distributed cloud of points now? (2-3 sentences.)  

### Answer 
### scale_x_log10() transforms the x axis scals into a log scale: base 10. In this case, the x axis is represented by 1000, 10 000, and 100 000 (1e+03, 1e+04, and 1e+05 in scientific notation). As the initial data was spread, the log function made it less spread and more evenly distributed. 

Nice! The x-axis now has scientific notation, let's change that.

**Exercise:** What does the `dollar()` call do?  

### Answer:
### it renamed the x variables from scientific notation to the dollar sign, which is more readable

```{r, dollar_answer}
?dollar()
```

**Exercise:** How can you find other ways of relabeling the scales when using `scale_x_log10()`?  

### Answer: 
###  we can use scale_x_log10(labels=scales::comma) instead of scale_x_log10(labels=scales::dollar) 

### The Ggplot Recipe

1. Tell the `ggplot()` function what our data is.
2. Tell `ggplot()` *what* relationships we want to see. For convenience we will put the results of the first two steps in an object called `p`.
3. Tell `ggplot` *how* we want to see the relationships in our data.
4. Layer on geoms as needed, by adding them on the `p` object one at a time.
5. Use some additional functions to adjust scales, labels, tickmarks, titles. 
  - The `scale_`, `labs()`, and `guides()` functions

### Answer:
### we can use labs() to add descriptions, e.g.:  
### labs(x = "GDP Per Capita",
###       y = "Life Expectancy in Years",
###       title = "Economic Growth and Life Expectancy",
###       subtitle = "Data points are country-years",
###       caption = "Source:Gapminder")


### Mapping Aesthetics vs Setting them

```{r, mapping_vs_setting}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = 'yellow'))
p + geom_point() + scale_x_log10()
```

This is interesting (or annoying): the points are not yellow. How can we tell ggplot to draw yellow points?

**Exercise:** Based on the discussion in Chapter 3 of *Data Visualization* (read it), describe in your words what is going on. 
One way to avoid such mistakes is to read arguments inside `aes(<property> = <variable>)`as *the property <property> in the graph is determined by the data in <variable>*. 

### Answer: 
### In the code above, the color is specified in the aes() function where that mapping is specified. So, the program is trying to map variable to the color aesthetic, assuming we're giving it a variable. It is not "yellow" because the default first-category hue is red. Therefore, the aes() function is for mappings only. 
### To set a property, we have to use geom_, e.g. p+geom_point(color = "yellow") 

**Exercise:** Write the above sentence for the original call `aes(x = gdpPercap, y = lifeExp, color = 'yellow')`.

### Answer: didn;t understand what "above sentence is". Assumed ti put color in geom_. The code is as follows:

```{r, changing_the_color}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(color = 'yellow') + scale_x_log10()
```


Aesthetics convey information about a variable in the dataset, whereas setting the color of all points to yellow conveys no information about the dataset - it changes the appearance of the plot in a way that is independent of the underlying data. 

Remember: `color = 'yellow'` and `aes(color = 'yellow')` are very different, and the second makes usually no sense, as `'yellow'` is treated as *data*.

```{r, exercise_args_for_smooth}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point() + geom_smooth(color = "orange", se = FALSE, size = 3, method = "lm") + scale_x_log10()
```

**Exercise:** Write down what all those arguments in `geom_smooth(...)` do.  

### Answer:
### color = "orange" - sets the line color to orange; 
### se = FALSE -  a ribbon showing the standard error for the line has been removed from its default value of TRUE to FALSE
### size = sets the line's size, or its thickness, to 8
### method = "lm" - lm stands for linear model, i.e. the orange line that we see in the plot

```{r, gapminder_with_labels}
p + geom_point(alpha = 0.3) + 
  geom_smooth(method = "gam") + 
  scale_x_log10(labels = scales::dollar) +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
       title = "Economic Growth and Life Expectancy",
       subtitle = "Data Points are country-years",
       caption = "Source: Gapminder")
```

Coloring by continent:

```{r, gapminder_color_by_continent}
library(scales)
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = continent, fill = continent))
p + geom_point()
p + geom_point() + scale_x_log10(labels = dollar)
p + geom_point() + scale_x_log10(labels = dollar) + geom_smooth()
```

**Exercise:** What does `fill = continent` do? What do you think about the match of colors between lines and error bands?  

### Answer:
### The fill aesthetic sets the color of the standard error ribbon, in this case, the interior of the smoother’s standard error ribbon.

```{r, gapminder_color_by_continent_single_line}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + geom_smooth() + scale_x_log10()
```

**Exercise:** Notice how the above code leads to a single smooth line, not one per continent. Why?  

### Answer:
### By default, geom_ inherit their mappings from the ggplot() function. We can change this by mapping the aesthetics. Using mapping = aes(...) in geom_, we specify only the mappings we want to apply.

**Exercise:** What is bad about the following example, assuming the graph is the one we want? This is why you should set aesthetics at the top level rather than at the individual geometry level if that's your intent.  

### Answer: 
### The plot below is very hard to read as it's very "noisy"


```{r, many_continents}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + 
  geom_smooth(mapping = aes(color = continent, fill = continent)) + 
  scale_x_log10() + 
  geom_smooth(mapping = aes(color = continent), method = "gam")
```

## Additional (not Optional) Exercises 

**Exercise (Discourse):** Find ways to save the figures that you made so that you can use them elsewhere too. Create a new folder to save only images. Use the command for saving to save the picture for the last image in your new folder, after you have updated the axes, title, subtitle, and caption of the image. Post your solution on Discourse and use it to include the final image above with a caption saying "Saved by <your-name>" inside your Discourse post.

### Answer:

```{r, additional_exercise1, fig.width = 12, fig.height = 9}

p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp))

p + geom_point(mapping = aes(color = continent, alpha=0.5)) +
    geom_smooth(mapping = aes(color = continent, fill = continent)) +
    scale_x_log10(labels=scales::dollar)+
    labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
         title = "Economic Growth and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Saved by Akylbek Subanbekov")

ggsave(filename = "Additional_exercise1.png")
```

**Exercise:** Read section 3.8 "Where to go next" from DV. Based on those ideas, experiment and create two different graphs with the gapminder data. Describe each briefly in one sentence.

###Answer:
### 2.1 change x and y, and reversed the graph, so it's upside down now
```{r, additional_exercise2.1, fig.width = 12, fig.height = 9}

rev <- ggplot(data=gapminder,
            mapping=aes(y=gdpPercap,
                        x=lifeExp))

rev+geom_point()+geom_smooth(method="gam")+
  scale_y_reverse(labels=scales::dollar)
```

### 2.2 the second variable becomes the population. The graph is trying to link life expectancy with population

```{r, additional_exercise2.2, fig.width = 12, fig.height = 9}

ppl <- ggplot(data=gapminder,
            mapping=aes(x=pop,
                        y=lifeExp))
ppl+geom_point(mapping=aes(color=continent))+geom_smooth(method="gam")+
  scale_x_log10(labels=scales::comma)
```
### 2.3 Just like the continents were colored previously, the graph now colors the years of the data, and it doesn't seem to make much sense

```{r, additional_exercise2.3, fig.width = 12, fig.height = 9}

yr <- ggplot(data=gapminder,
               mapping=aes(x=gdpPercap,
                           y=lifeExp,
                           color=year,
                           fill=year))
yr+geom_point()+
  geom_smooth(method="loess")+
  scale_x_log10()

```


**Exercise:** Read section 1.6 of [R for Data Science](https://r4ds.had.co.nz/introduction.html) on *Getting help and learning more*. Go back to an error from your previous assignment -- or pick a new one -- and post a reproducible error as described in that section on the discourse forum.

## Done [link](https://discourse.trichotomy.xyz/t/assignment-3-getting-help-and-posting-errors/73/5?u=akylsu)



**Exercise:** Do exercise 3.2.4 from [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html#first-steps). Include your code in chunks, describe the output and code (where necessary) in the surrounding text.

## Answer:

```{r, exercise3.2.4.1}

#Run ggplot(data = mpg). What do you see?

ggplot(data = mpg)

# I see a blank grey rectangle

```

```{r, exercise3.2.4.2}
#How many rows are in mpg? How many columns?

mpg

# There are 11 columns and 234 rows

```

```{r, exercise3.2.4.3}
#What does the drv variable describe? Read the help for ?mpg to find out.

?mpg

# f = front-wheel drive, r = rear wheel drive, 4 = 4wd

```

```{r, exercise3.2.4.4}
#Make a scatterplot of hwy vs cyl.

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy))

# it seems the fewer cylinders a car has, the less fuel it consumes

```

```{r, exercise3.2.4.5}

# What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

?mpg

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))

# "class" is a "type" of a car; "drv" is front wheel, rear wheel, or 4wd 
# it shows the drive and the class of a car. It seems that this graph is meaningless becauses a certain class of a car can be front, rear, or 4wd

```

?mpg

**Exercise:** Go through Exercises in 3.3.1. If an exercise does not make sense immediately (such as you don't know what a categorical variable is), replace the question by a question that addresses that point (in the case of the caregorical variable "What are categorical and continuous variables and how are they different in R?"). Write it down, try to answer that question, and ignore the original question. That way you don't end up spending too much time on this one exercise.

## Answer:

```{r, exercise3.3.1.1}
#1. What’s gone wrong with this code? Why are the points not blue?
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))

# As dicsussed earlier, it's not blue because geom_point(mapping = aes(...)) treats the color as a variable.
```

```{r, exercise3.3.1.2}
#2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?
?mpg
#Note: In a categorical variable, the value is limited and usually based on a particular finite group. For example, a categorical variable can be countries, year, gender, occupation. A continuous variable, however, can take any values, from integer to decimal.

##Categorical: manufaturer and model (there is a certain number of manufaturers and models at a given period of time), year, drive, cylinder, trans, fl (fuel type), class
##continuous: cty, hwy, 
```

```{r, exercise3.3.1.3}
#3.1 Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = hwy))

## the more fuel efficient the car is, the lighter is a color of a dot

```
```{r, exercise3.3.1.3.2}
## 3.2 Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

#ggplot(data = mpg) + 
#  geom_point(mapping = aes(x = displ, y = hwy, shape = hwy))

## Error: A continuous variable can not be mapped to shape

```

```{r, exercise3.3.1.3.3}
#3.3 Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = hwy))

## the more fuel efficient the car is, the larger is a dot

```


```{r, exercise3.3.1.4}
#4 What happens if you map the same variable to multiple aesthetics?

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = hwy, size = hwy))



## in this case, a variable hwy was assigned to 3 different aesthetics. As you can see, it was assigned to y axis with both color and size

```


```{r, exercise3.3.1.5}
#5 What does the stroke aesthetic do? What shapes does it work with? 

ppl <- ggplot(data=gapminder,
            mapping=aes(x=pop,
                        y=lifeExp))
ppl+geom_point(mapping=aes(color=continent))+geom_smooth(method="gam")+
  scale_x_log10(labels=scales::comma)

car <- ggplot(data = mpg,
              mapping= aes(x=displ,y=cty))
car + geom_point(shape=21, colour = "orange", fill = "black", size = 10, stroke = 2)

# stroke aesthetic is to modify the width of the border, works with area and as such

```

```{r, exercise3.3.1.6}
#6 What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

car6 <- ggplot(data = mpg,
              mapping= aes(x=displ,y=cty, colour = displ < 5))
car6 + geom_point()

# it colored the engine displacements that are less than specified amount(in this case 5)



```

**Exercise:** Read the (very short) [Chapter 4 of R for Data Science](https://r4ds.had.co.nz/workflow-basics.html) and try exercise 1 in section 4.4.

```{r, exercise4.4.1}
# Why does this code not work?


my_variable <- 10
my_varıable

# it works perfectly for me. Yet,  I guess there is a typo - using in "i" in the 1st expression and "ı" in the 2nd one

```


```{r, exercise4.4.2}
# Why does this code not work?

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8)
#filter(diamond, carat > 3) # no such variables


```

**Bonus Exercise:** Why did I load the `scales` library twice via `library(scales)` to knit? 

## Because in the first instance, the {eval=FALSE}, so the code will not be evaluated and its results will not be included in the knitted doc

install.packages('tinytex')
tinytex::install_tinytex()

## Assignment 3

1. Do the exercises in these lecture notes. 
#Done
1. Knit lectures 2, making sure to get rid of those `eval=FALSE` that are just there because I didn't complete the code
#Note: when I removed 'eval+FALSE', the document wouldn't knit. So, assumed I have to completely remove the incomplete code
1. Upload your pdf on Moodle
#Done
1. Grade assignment 2 on Moodle -- let me know if you can't access Moodle!
#Done
1. If you are part of the team that does the first group assignment, start thinking about how you are going to do the assignment. You have until lecture 4.
#na