---
title: "Lecture 12"
author: "Marc Kaufmann"
date: "12/2/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Other Features of R

### `map` and friends

Some quick examples.

See [https://r4ds.had.co.nz/iteration.html#the-map-functions] for some examples on using map and [https://r4ds.had.co.nz/iteration.html#other-patterns-of-for-loops] for `reduce`. These are common functions in particular in functional programming languages, although there is nothing disallowing them in other languages. They are great replacements for various types of for-loops and iterations, although it may take some time to get used to them. Once you do get used to them, they can be quite succinct, and work well in pipes.

### Sourcing code

You can load other code via `source(...)`, e.g. another .R file that has some tools

```{r}
source('toy_tools.R')
sourcerer()
```

This is most useful for some utilities that you keep using, or some such. 

### Connecting directly to databases

When your dataset gets large, it can be more efficient (or necessary) to keep it in a database. R allows you to connect to the database and perform actions there. Once you have selected (in the database) the subset of data you want to work on, you can fully load it into R to perform analysis on it.

```{r include=FALSE}
library(RSQLite)

# Create toy in-memory database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
summary(con)
dbListTables(con)
dbWriteTable(con, "flights", nycflights13::flights)
dbListTables(con)

# Can do standard SQL
dbListFields(con, "flights")
dbReadTable(con, "flights") 
res <- dbSendQuery(con, "SELECT year, month, day, dep_time, arr_delay FROM flights WHERE arr_delay > 120")
dbFetch(res)
dbClearResult(res)
dbDisconnect(con)
res
# Or do simpler with R
```

```{r, database_again, include=FALSE}

# Or use dplyr from tidyverse
library(tidyverse)
con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbListTables(con)
copy_to(con, nycflights13::flights, "flights")
flights_db <- tbl(con, "flights")
dbListTables(con)
tailnum_delay_db <- flights_db %>% 
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay),
    n = n()
  ) %>% 
  arrange(desc(delay)) %>%
  filter(n > 100)

tailnum_delay_db
tailnum_delay_db %>% show_query()
# we did not yet get all the data
nrow(tailnum_delay_db)
tail(tailnum_delay_db)
# Once you got the query right, use collect() to get all the data
tailnum <- tailnum_delay_db %>% collect()
dbDisconnect(con)
tailnum
tailnum_delay_db
```


### Faster/more efficient code

```{r, for_loops}
mean_of_squares <- function(n) {
  s <- 0
  for(i in seq_len(n)) {
    s <- s + (i*i/n)
  }
  s
}

mean_of_squares(2)
mean_of_squares(3)
system.time(mean_of_squares(40000))

vec_mos <- function(n) {
  mean((1:n)^2)
}

system.time(vec_mos(40000))

```

Another classic:

```{r, fibonacci}

fibo <- function(n) {
  if (n == 1 | n == 2) {
    1
  } else {
    fibo(n-1) + fibo(n-2)
  }    
}
fibo(5)
fibo(6)
fibo(7)

system.time(n <- fibo(31))
n
#system.time(n <- fibo(32))
#n

smart_fibo <- function(n, last_fibo, second_to_last_fibo) {
  if (n == 2) {
    return(last_fibo)
  } else {
    return(smart_fibo(n - 1, last_fibo + second_to_last_fibo, last_fibo))
  }
}

smart_fibo_wrap <- function(n) {
  if (n > 2) {
    smart_fibo(n, 1, 1)
  } else {
    1
  }
}

smart_fibo_wrap(5)
smart_fibo_wrap(8)
smart_fibo_wrap(32)
smart_fibo_wrap(40)
smart_fibo_wrap(100)
system.time(smart_fibo_wrap(32))
```

#### Data Table

If you need more performance, you should use `data.table`. It has a [useful vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html). 

## Other Features of RStudio

There are several packages that help making books, articles, or blogs:

```{r gratuitous_define}
too_much_time <- 3
```

- `bookdown` for books, which can be used for making articles too
  - I spent `r too_much_time`+ hours at some point fiddling around with it to get it to work properly with articles. Oh the fun...
- `blogdown` for blogs
- Slides for presentations

### Include Python Code

```{r}
library(reticulate)
```

```{python}
def square(x):
    return(x*x)

square(1)
square(-2)
```

```{r, eval=FALSE}
res <- py$square(3)
```

With the library `reticulate`, you can even call python things from R code.


```{python}
def square(x):
    return(x*x)

square(1)
square(-2)
```

```{r}
py$square(3)
```

For more, see [https://rstudio.github.io/reticulate/articles/r_markdown.html]. Without it, you can still add Python code, but you cannot share objects between your Python and R code.

## Where Next?

One advantage of RStudio and R tools is that most of what we do here can be done on the command line too. Thus you can (and I do) integrate these tools into shell scripts or make files or the like. This has two major benefits:

1. You can script tasks, whether simply for convenience or because you want to automate it as part of a pipeline
2. You can move away from RStudio, while keeping much of the workflow.
  - I don't like editing code in RStudio when I don't need to explore data or visualizations.
  
Kieran Healy has another book at [http://plain-text.co/] on how to work with text files. It is geared towards academics since it focuses on replicability, but much of it should apply to data analysts and scientists.

R4DS has a nice final page that covers a few of the above and some more: [https://r4ds.had.co.nz/r-markdown-formats.html]. In particular, it mentions dashboards as a separate content type which I hadn't known about. It looks nifty.

Finally, the material covered in this class can be used relatively straightforwardly in Python and Julia:

- Python: 
  - dplyr (+ other parts of tidyverse) are similar to pandas (+ ??)
  - ggplot2: plotnine package seems similar; matplotlib seems different
- Julia:
  - Dataframes.jl is similar to dplyr
  - Gadfly.jl is similar to ggplot2
  
See https://dataframes.juliadata.org/stable/man/comparisons/ on a comparison from the Julia side. As you'll see, much of what you learned can be used with small syntactic changes. In fact, you can call R and Python code from Julia, which is helpful when there are missing packages in Julia, which is the youngest language.