---
title: "Lecture 12"
author: "Marc Kaufmann"
date: "12/2/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Other Features of R

### `map` and friends

Some quick examples.

### Data Table

If you need more performance, you should use `data.table`. It has a [useful vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html). 

### Sourcing code

```{r}
source(...)
```

### Connecting directly to databases

...

## Other Features of RStudio

There are several packages that help making books, articles, or blogs:

- `bookdown` for books, which can be used for making articles too
  - I spent 3+ hours last week fiddling around with it to get it to work properly with articles. Oh the fun...
- `blogdown` for blogs
- Slides for presentations

### Includ Python Code

```{python}
def square(x):
    return(x*x)

square(1)
square(-2)
```

```{r, eval=FALSE}
py$square(3)
```

With the library `reticulate`, you can even call python things from R code.

```{r}
library(reticulate)
```

```{python}
def square(x):
    return(x*x)

square(1)
square(-2)
```

```{r}
py$square(3)
```

For more, see [https://rstudio.github.io/reticulate/articles/r_markdown.html].

### Shiny Apps

...

## Where Next?

One advantage of RStudio and R tools is that most of what we do here can be done on the command line too. Thus you can (and I do) integrate these tools into shell scripts or make files or the like. This has two major benefits:

1. You can script tasks, whether simply for convenience or because you want to automate it as part of a pipeline
2. You can move away from RStudio, while keeping much of the workflow.
  - I don't like editing code in RStudio when I don't need to explore data or visualizations.
  
Kieran Healy has another book at [http://plain-text.co/] on how to work with text files. It is geared towards academics since it focuses on replicability, but much of it should apply to data analysts and scientists.

R4DS has a nice final page that covers a few of the above and some more: [https://r4ds.had.co.nz/r-markdown-formats.html]. In particular, it mentions dashboards as a separate content type which I hadn't known about. It looks nifty.