---
title: "Materials for Class"
author: "Marc Kaufmann"
date: "August 29, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Error in library(reticulate) : there is no package called 'reticulate'
Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> library
Execution halted

You can even run python scripts in RMarkdown if you have everything necessary installed on your computer.

```{r}
knitr::opts_chunk$set(engine.path = list(python = '/usr/bin/python'))
install.packages("reticulate")
library(reticulate)
```

And then do

```{python}
def square(x):
  return x*x
  
square(11)
```

day1 <- day1 %>% 
  select(day1, c(X = answer_1, Xprime = answer_2, Y = answer_3, Yprime = answer_4)) %>%
  mutate(deltaX = Xprime - X, deltaY = Yprime - Y) %>%
  mutate(deltadelta = deltaY - deltaX)

Beautiful error:
Error: `day1` must evaluate to column positions or names, not a list Call `rlang::last_error()` to see a backtrace

When passing a dataframe with one column into t.test, I get: "Error: Must use a vector in `[`, not an object of class matrix. Call `rlang::last_error()` to see a backtrace"


Did you define a new function in another chunk? Did you forget to run the other chunk? Welcome to 

"Error in function_list[[i]](value) : could not find function "get_batch2_later_sessions""

