---
title: "Lecture 1"
author: "Marc Kaufmann"
date: "9/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This introduction is essentially Chapter 2 from Kieran Healy's book [Data Visualization](http://socviz.co/), which you should read as part of Assignment 2 and to find your way around RStudio. 

We will use exclusively R Markdown files. In this case, you should simply open the file 'lecture1.Rmd' from the repository and you are set. Whenever you create a new file, you should choose to make an R Markdown file.

### Things to Know about R

There are 4 things to bear in mind about R:

1. Everything has a name
2. Everything is an object
3. You do things using functions
4. Functions come in packages

#### Everything has a name

Everything that you use in R has a name: variables (including datasets), functions, or special reserved words. That is the way you talk about them. Here are some examples:

```{r, eval=FALSE}
# Numbers are called by their number, arithmetic operations by their usual symbol
3
4 + 7
2 * 3
7/3
7 %% 3

# There are some pre-defined variables
pi
"pi"
'pi'
TRUE
True
true

# There are also pre-defined functions
c
"c"
"C"
```

**Some Notes**:

- '#' is the comment sign. It tells R that everything that follows is a comment and R should ignore it. Comments are there to help explain parts of the code that need additional documentation. We will overuse them initially, but reduce this as the course proceeds.
- c, "c", and "C" are not the same things
- Naming conventions: When you name variables and functions, you should use snake case.

**Exercise:** For each of the lines in the following code chunk, write in a comment next to it what it does. I completed the first example for you.

```{r, exercise1, eval=FALSE}
3         # -> 3
7 %% 3    # -> ?
False     # -> ?
'pi'      # -> ?
FALSE     # -> ?
```

**Exercise:** Look up what 'snake case' is and add your answer here.

**Exercise:** What types of names are allowed in R? Look it up and write your answer here as you understand it. Then provide 3 examples of things that are not valid names in R for different reasons. 

#### Everything is an object

There are built-in objects and objects you import or create. Most importantly, you assign values to objects you create with the '<-' operator: a '<' ('less than') followed by a '-' ('minus' or 'dash').

```{r, assignment}
marcs_new_object <- "I am a fancy object"
marcs_new_object
```

One important function to create objects is the $c()$ function, which combines several items into one objects:

```{r}
marcs_new_combined_object <- c(marcs_new_object, "Some apples, because they are healthy")
marcs_new_combined_object
```

In principle, you can use '=' to assign values, but this is very non-idiomatic R. In this course, I will subtract points for using '='. Use '<-'.

**Class Exercise:** How many of the following programming languages have you *heard* of? 

- R
- SQL
- Racket
- Lisp
- JavaScript
- ECMAScript
- bash
- C
- Perl
- Logo
- Scratch

```{r}
languages_heard_of <- c(11)
```

#### You do things using functions

> A function is a thing that does things to things. 

A function is a special object that you can call, say the function $mean$. It is an object:

```{r}
mean
```

but as a simple object, it is a bit boring. When we *call* it, we put parentheses at the end, and we tell it what to perform its action on:

```{r}
mean(c(0,10))
mean(languages_heard_of)
```

**Exercise:** What happens if you call the function $mean$ without any arguments, i.e. $mean()$?

#### Functions come in packages

As you start seeing, lots of functions are already defined for you, such as $mean()$ and $sd()$ and $c()$. There are many more functions that were written by other people and bundled into packages so that you can use them. Many packages come bundled with R from the start (base R), others you can install on your system via 'install.packages()' and use via 'libarry()'. We will soon use the 'ggplot()' function. Try using it now:

```{r, eval=FALSE}
ggplot
```
You will get an error. We first need to import the library:

```{r}
library(ggplot2)
ggplot()
```

This gives us a beautifully empty plot. 


### How to figure out what is what

Suppose you have an object $x$ that you don't know what it is. You can do a few things to find out:

```{r, what_is_x}
x <- c(1, 3.0, 2.9)
x
str(x)
class(x)
```

For built-in functions, you can also ask for help or bring up the documentation with 'help()' or '?' or '??':

```{r, what_is_summary}
help(summary)
?summary
?class()
?str()
```

**Exercise:** What do 'class()' and 'str()' do? Use 'help' (or '?'). Don't spend too much time reading the docs. Which description do you find more helpful?

### Make Your First Figure

```{r, gapminder}
library(gapminder)
gapminder
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()
```

## Assignment 2

This assignment is due before the start of class 2. Commit your knitted assignment2.html to your git and push it to your repository regularly as you make progress.

- Only your work until 13:30 will be considered. 
- If you know that you might miss the deadline, you have to email me in advance and I will tell you whether you can get an extension. No extension will be granted less than 24 hours prior to the deadline -- you should have started working on it.
- If you miss the deadline (even by 1 minute) you lose 25% irrespective of technical issues such as 'the internet went down'. It wasn't down the whole week, and you knew a deadline was coming. In a business setting, you would probably lose more than 25%. Repeat offenders will face an increasingly large penalty.
- If you struggle *answering* the assignment, you should submit what you tried, and send me a message that you struggled. Much of the grade initially is on trying out things, even if it doesn't work out, so you should submit.

### Part 1

Read chapter 2 of [Data Visualization](http://socviz.co/gettingstarted.html). Make sure you know where the console is, the editor, and what code chunks are in RStudio. 

### Part 2

Complete all the exercises in the lectures notes of lecture 1. Put your answer to an exercise right after the question of the exrcise in the R Markdown file. When done, knit the document, commit the changes with the commit message "Part 1 of assignment 2" and push them to your GitHub repository.

### Part 3

Fill in the holes in the lecture notes of lecture 1. When done, knit the document, commit the changes with the commit message "Part 3 of assignment 2" and push them to your GitHub repository.

**Note:** If you can't figure out how to use git, post a question on the discourse forum and send me the html file by email.

### Part 4

Track as many error messages as possible that you made during this week (stop once you get to 4). Put the code that caused the error in the chunk below, and copy the error message as a comment below it. I provide an example.

```{r, eval=FALSE}

# Example: "4" - "3" leads to the error 'Error in "4" - "3" : non-numeric argument to binary operator'
# Thus you write:
"4" - "3"
# -> Error in "4" - "3" : non-numeric argument to binary operator

# Your examples below
```

Post one of these 4 errors on discourse.

## Resources for this lecture 