---
title: "Intro to R: Lecture 1"
author: |
  | Marc Kaufmann
  | Central European University
format: 
  beamer:
    theme: metropolis
    slide-level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction to Data Analysis in R

## Who am I?

\pause{}

- Assistant Professor in Economics and Business
- Research in Psychology and Economics (aka Behavioral Economics)

\pause{}

Relevant for this class:

- Collect my own data in (mostly online) experiments
- Analyze said data: Fairly basic, since design is up to me 
- Program in Racket, R, bash/unix, and Python (in that order of competence)

\pause{}

Most importantly: I am good at getting help.

## Goal for the class in one sentence

\pause{}

> Generate basic insights\pause{} from existing data\pause{} that is small\pause{} and relational\pause{} in a reproducible and replicable manner.

\pause{}

## Goals of the Course

```{r, eval=FALSE, echo=TRUE}
code_1 <- function(collected_data) { 
  collected_data |>
    read()       |>
    clean()      |>
    explore()    |>
    visualize()  |>
    summarize()  |>
    tidy()       |>
    analyze()    |> 
    render()
}
```

## Focus of the Class

Since we cover much of the data analysis cycle and have little time:

- Focus on few libraries and commands (80/20 rule)
  - Tidyverse only: coherent set of tools with sane interface
  - Has inspired other languages (Julia, Python)
- Focus on *fluency* (rather than complete understanding)
- Focus on teaching you how to learn
  - Write readable, self-documenting code
  - How to get help

## Assignments and Grading

- Data Reports (100 points): 4 short reports with simple data
  - DR1: 15 points
  - DR2: 20 points
  - DR3: 25 points
  - DR4: 40 points
- Exercises (0 points, for practice)
  
## Data Report

Rubric for data report written in Quarto (.qmd, format: html):

- Describe dataset: 20%
  - Where does it come from, what do the variables mean? You may have to ask.
- Data cleaning: 20%
  - Always keep the raw data, create new files for the cleaned data
  - Explore and describe cleaning process, provide summaries (tables, summary stats)
- Explore for insight: 40%
  - Include one visualization and some analysis (summaries, mostly) and explain
- Write up and conclusions: 20%

Report has to be replicable and well documented (name, etc). 

## Data Reports: Tentative Deadlines

Most likely schedule (depends somewhat on progress):

- Data Report 1: week 4 (deadline week 5)
- Data Report 2: week 6 (deadline week 7)
- Data Report 3: week 8 (deadline week 9)
- Data Report 4: week 10 (deadline week 12)

You lose 33% of score for how each day late (rounded up). No ex-post exceptions.

## Useful Resources

When you get stuck, the following fantastic books may help:

- Kieran Healy's book on *Data Visualization* (http://socviz.co/)
- Grolemund and Wickham's book *R for Data Science* (https://r4ds.had.co.nz/)

Additionally:

- Ask questions on Slack

## Where you should be

You should:

- Have RStudio set up up and working
- Have git set up and working
- Have joined Slack and posted your intro

## Questions?

\pause{}

Any questions?
