---
title: 'Coding 1: Lecture 1'
author: |
  | Marc Kaufmann
  | Central European University
date: "9/10/2019"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation:
    latex_engine: xelatex
    slide_level: 2
    theme: metropolis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Coding 1: Data Management and Analysis with R

## Basic Admin

- Weeks 1 to 6: Mondays 13:30-15:10
- Weeks 7 to 12: Mondays 17:30-19:10
- Instructor: Marc Kaufmann (call me Marc)
- Teaching Assistant: Júlia Hermann gives 3 sessions
  - 17:30-19:10 on Tuesday October 8th
  - 17:30-19:10 on Tuesday November 5th
  - 17:30-19:10 on Tuesday November 19th

## Who am I?

\pause{}

- Assistant Professor in Economics and Business
- Research in Psychology and Economics (aka Behavioral Economics)

\pause{}

Relevant for this class:

- Collect my own data in (mostly online) experiments
- Analyze said data: Fairly basic, since design is up to me 
- Program in Racket, bash/unix, Python, and R (in that order of competence)

\pause{}

Most importantly: I am good at getting help.

## Goal for the class in one sentence

\pause{}

> Generate basic insights\pause{} from existing data\pause{} that is small\pause{} and relational\pause{} in a reproducible and replicable manner.

\pause{}

Or expressed in R (pseudo-)code:

\pause{}

```{r, eval=FALSE, echo=TRUE}
library(msc_ba)

eureka_or_bust <- your_great_ideas %>% 
  # We need data to figure this out. Let's...
  collect_data() %>%
  # Apply knowledge from this class...
  code_1()
```

## Goals of the Course

```{r, eval=FALSE, echo=TRUE}
code_1 <- function(collected_data) { 
  collected_data %>% 
    read_in() %>%
    explore() %>%
    visualize() %>%
    summarize() %>%
    clean() %>%
    tidy() %>%
    analyze() %>%
    knit()
}
```

```{r, eval=FALSE}
  # The data is now in a data base. Or in a .csv file. So let's...
  read_the_data_into_R() %>%
  # Total insight soon. Let's...
  look_at_the_data() %>%
  # Ergh, some of the data got miscoded. So let's...
  clean_the_data() %>%
  # Is it better? Let's...
  look_at_the_data_again() %>%
  # Much better. But data is not tidy. Let's...
  tidy_the_data() %>%
  # Ready to analyze! Surely everyone loves my idea, I'll be rich!!
  analyze_data_for_insight() %>%
  # The fools don't love my idea! I'd better...
  create_report_for_disappointed_supervisors() 
```

## Focus of the Class

Since we cover much of the data analysis cycle and have little time:

- Focus on few libraries and commands (80/20 rule)
  - Tidyverse only: coherent set of tools with sane interface
- Focus on correct code; then maintainable; then fast
- Focus on *fluency* 
- Focus on teaching you how to learn
  - Clean code; documenting; debugging; communicating
  - How to get help

## Focus of the Class

> Strive to write code that is correct; maintainable; and fast. The ordering of these adjectives is critical: correct is more important than maintainable; maintainable is more important than fast; and fast is important to include, because nobody wants to live with slow programs. 
>
>
> From "How to Program Racket: a Style Guide", Felleisen et al

## Assignments and Grading


- Participation (30%): helping yourself and helping others 
  - Includes attendance. Starting week 2, will take attendance via https://www.youhere.org/
    - Let me know if you cannot or do not want to use that.
- Assignments (77-78%): weekly assignments in the form of R Markdown notebooks, and assessing those of your peers
  - Grading is N% in week N ($N = 1, ..., 12$)
- No exam
  
\pause{}

Total: 107-108%

## Useful Resources

When you get stuck, the following fantastic books may help:

- Kieran Healy's book on *Data Visualization* (http://socviz.co/)
- Grolemund and Wickham's book *R for Data Science* (https://r4ds.had.co.nz/)

Additionally:

- Tuesdays 17:30-19:10, N13 309: Coding practice with R
- Júlia's sessions
- Ask questions on https://discourse.trichotomy.xyz

## Where you should be

You should:

- Have RStudio set up up and working
- Have git set up and working
- Have cloned the repository for the course
- Have an account on https://discourse.trichotomy.xyz

## First time I teach R

I tend to experiment quite a bit:

- Early on, still worth figuring out what works
- Especially assignments and how to incentivize teamwork
- After week 3, should be ironed out

\pause{}

Any questions?