---
title: "Lecture 6"
author: "Marc Kaufmann"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Class Exercises

The following chunks have most, but not all of the code necessary to achieve their goal. Fix as many of them as you can. **Note:** You'll have to switch off the `eval = FALSE`.

```{r, eval = TRUE}
# Compute for every day how many minutes above the average delay a given flight is.

# 1. Compute the average delay (per day) and put it in the rows
# 2. Then use that with dep_delay to compute the minutes above average
library(tidyverse)
library(nycflights13)

flights_small <- flights

flights_small %>%
  group_by(year, month, day) %>%
  mutate(difference_from_daily_mean_dep_delay = dep_delay - mean(dep_delay, na.rm=TRUE)) %>%
  select(difference_from_daily_mean_dep_delay, everything())
```

Now compute the daily standard deviation in departure delays. Explain in one sentence what it means for that to be higher or lower.

```{r}
# Compute for every day how many minutes above the average delay a given flight is.

flights_small %>%
  group_by(year, month, day) %>%
  mutate(daily_sd_dep_delay = sd(dep_delay, na.rm=TRUE)) %>%
  select(daily_sd_dep_delay, everything())
```

Now combine the previous two commands with a third to compute how many standard deviations a given flight is from the mean. A value of $-1$ indicates that the flight had 1 sd less departure delay than the average flights that day, $1$ indicates that it had 1 sd more departure delay than the average flight that day.
  
```{r, eval = FALSE}

flights_small %>%
  group_by(year, month, day) %>%
  filter(!is.na(dep_delay)) %>%
  mutate(
    diff_mean_in_sd = (dep_delay - mean(dep_delay)) / sd(dep_delay)
  ) %>%
  select(diff_mean_in_sd, dep_delay, everything())
```
    
And finally (if time permits, which is unlikely):

```{r, eval = FALSE}
# What time of the day should you fly to avoid delays the most?
# Start with dep_time. Then realize this is bad.

not_missing %>%
  mutate(hour = dep_time %/% 100) %>%
  group_by(hour) %>%
  summarise(delay = mean(arr_delay))
  
# Fix the variable, check what it computes before ranking
not_missing %>%
  mutate(hour = ...%/% 100) %>%
  group_by(hour) %>%
  summarise(delay = mean(arr_delay))

# Now rank (in reality you would change the earlier code, rather than repeat)
not_missing %>%
  mutate(hour = ...%/% 100) %>%
  group_by(hour) %>%
  summarise(delay = mean(arr_delay)) %>%
  mutate(rank_delay = rank(delay)) %>%
  arrange(rank_delay)
```

## Ceci n'est pas a tibble

If you have a data frame that is not a tibble you can change it to one with `as_tibble()`:

```{r}
ceci_nest_pas_a_tibble <- iris
is_tibble(ceci_nest_pas_a_tibble)
ceci_nest_pas_a_tibble # Run in console, rstudio seems to do conversion automatically
a_tibble <- as_tibble(ceci_nest_pas_a_tibble)
a_tibble
```

Sometimes we want to call a function at the end of a pipe that does not take a data frame, but a vector. For instance if we want to compute a t-test:

```{r}
df <- tibble(
  x = runif(100),
  y = runif(100, min=-0.5, max=0.5),
  z = rnorm(100)
)

df %>% select(x) %>% t.test() # ERROR!
df %>% .$x # OK
df %>% t.test(.$x) # ERROR?! WHAT? Somewhat baffled.
df %>% .$x %>% t.test() # A bit tedious
df %>% select(x) %>% deframe() %>% t.test() # Just to show that we can be even more tedious

# And if you can't figure it out... you don't have to pipe everything. 

t_test_this <- df$x
t.test(t_test_this)

# For t.test(x,y): I couldn't figure out one of the above ways.
t.test(df$x, df$y)
library(magrittr)
# Notice it is %$%, not %>%
df %$% t.test(x, y)  # Now we're talking


```

The above may not often be sufficiently justified to go through the hassle. Just assign to a variable and deal with vectors directly.

## Chapter 11: Data Import

Now on to importing data.

```{r}
library(tidyverse)

(test_data1 <- read_csv("test-data.csv"))

read_csv(
  "djflsjflkdsjfklsdjfkl
   a,b,c
   1,2,3
   4,5,6",
  skip = 1)

read_csv(
   "1,2,3
   4,.,6",
  col_names = FALSE,
  na = ".")

?read_delim
?read_tsv

# Class Exercise 5 in 11.2
# IDentify what is wrong with each of 

read_csv("a,b\n1,2,3\n4,5,6")
read_csv("a,b,c\n1,2\n1,2,3,4")
read_csv("a,b\n\"1")
read_csv("a,b\n1,2\na,b")
read_csv("a;b\n1;3")

# 11.3 Parsing a vector

str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-10-01", "1974-01-14")))

x <- parse_integer(c("123", "234", "abc", "12.3"))
x
problems(x)

parse_double("11.23")
parse_double("11,23")

parse_number("$100")

# 11.4

challenge <- read_csv(readr_example("challenge.csv"))
challenge
challenge <- read_csv(readr_example("challenge.csv"),
                      col_types = cols(
                        x = col_double(),
                        y = col_date()))

```


