---
title: "Assignment 4"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Follows Grolemund and Wickham, chapter 5


Install the dataset if you don't have it
install.packages("nycflights13")

```{r, eval = FALSE}
library(nycflights13)
flights
?flights
View(flights)
view(flights)
?view

library(tidyverse)
```

The following functions will be covered in this markdown:
1. filter()
2. arrange()
3. select()


Filtering (choosing) rows with filter()

```{r}
library(nycflights13)
library(tidyverse)
#filter(flights, month = 1) # Produces an error
filter(flights, month == 1)
filter(flights, month == 1, day == 1)
filter(flights, dep_time == 517)
```

dplyr functions don't change the data frame that you give it. They return a new one.
flights
Save the filtered data

```{r}
jan1 <- filter(flights, month == 1, day == 1)
jan1
```


Assign and print, use (varname <- ...)
```{r}

(feb1 <- filter(flights, month == 2, day == 1))
```

# Check it really assigned
```{r}

feb1
```


Some notes on comparisons

```{r}
sqrt(2)^2 == 2
sqrt(4)^2 == 4
(1/3)*3 == 1
1/49*49 == 1
1/(7^9)*7^9 == 1
```


In short, you can't rely on "It works because it works for what I tried".
For floating point comparisons, use near() to compare numbers

```{r}
near(sqrt(2)^2, 2)
```



Exercise: What counts as near? Find out. Can you change it?

Multiple constraints
|: is 'or' operator

```{r}
(jan_feb <- filter(flights, month == 1 | month == 2))
(not_jan <- filter(flights, !(month == 1)))
```



Class exercise: How do we know these actually worked?

```{r}

filter(not_jan, month == 1)
#View(jan_feb)
unique(not_jan$month)
jan <- filter(flights, month == 1)
nrow(flights) == nrow(jan) + nrow(not_jan)
(jan_to_june1 <- filter(flights, month <= 6))
jan_to_june2 <- filter(flights, month %in% c(1,2,3,4,5,6))
```


Check same number of observations

```{r}
nrow(jan_to_june1) == nrow(jan_to_june2)
```



Class Exercise: What does this do?

```{r}
mystery_filter <- filter(flights, !(arr_delay > 120 | dep_delay > 120))
mystery_filter2 <- filter(flights, arr_delay <= 120, dep_delay <= 120)
mystery_filter
mystery_filter2
identical(mystery_filter2, mystery_filter)
```



arr_delay > 120 OR dep_delay > 120
True if:
(140, 140)
(140, 0)
(0, 140)
 (0, 0)
Vote:
1. All flights that started and landed 120 minutes late
2. All flights that started 120 minutes late or landed 120 minutes late
3. All flights that started less than 120 minutes late or landed less than 120 minutes late
4. All flights that started and landed less than 120 minutes late

Class Exercise: Get the filter command for number 3 above
3. All flights that started less than 120 minutes late or landed less than 120 minutes late

```{r}
number3 <- filter(flights, arr_delay <= 120 | dep_delay <= 120)
number3 <- filter(flights, arr_delay < 120 | dep_delay < 120)
```




Class Exercise: get all flights that departed with less than 120 minutes delay but arrived with more than 120 minutes delay.

```{r}
dep_ok_arr_not <- filter(flights, dep_delay <= 120, arr_delay > 120)

ggplot(data = dep_ok_arr_not,
       mapping = aes(x = dep_delay)) +
  geom_histogram() + 
  labs(x = "Delay in departure time", y = "No. of Flights",
       title = "No. of Flights with Departure Delay =< 120 & Arrival Delay > 120",
       caption = "Source: nycflights")
```




Let's look at the data to see what the departure was for planes that arrived late but didn't start quite as late

```{r}
ggplot(data = flights,
       mapping = aes(x = dep_delay)) + 
  geom_histogram() + 
  labs(x = "Delay in departure time", y = "No. of Flights",
       title = "No. of Flights with Arrival Delay",
       caption = "Source: nycflights")
```



Filter flights by those that had dep_delay <= 120, then plot histogram

```{r}
dep_ok <- filter(flights, dep_delay <= 120)
ggplot(data = dep_ok,
       mapping = aes(x = dep_delay)) + 
  geom_histogram() +
  labs(x = "Delay in departure time", y = "No. of Flights",
       title = "No. of Flights with Dep Delay <= 120",
       caption = "Source: nycflights")
```




NA: Not available

```{r}
NA > 5
10 == NA
NA == NA
FALSE & NA
TRUE & NA
NA & FALSE
```



Nice example from G&W

Let x be Mary's age. We don't know how old she is.


```{r}
x <- NA
```

Let y be John's age. We don't know how old he is.

```{r}
y <- NA
```

Are John and Mary the same age?

```{r}
x == y
```

We don't know!


```{r}
NA^0
0 * NA
```


```{r}
is.na(x)
df <- tibble(x = c(1, NA, 3))
df
filter(df, x > 1)
filter(df, x > 1 | is.na(x))
```



```{r}
#arrange()

flights
arrange(flights, year, month, day)
arrange(flights, dep_delay)
arrange(flights, desc(dep_delay))

arrange(df, x)
arrange(df, desc(x))
```



Class exercise (do at home): How can we get the missing values at the top?

```{r}
arrange(flights, desc(is.na(arr_delay)))

```


Fastest flight

```{r}
colnames(flights)

arrange(flights, air_time)
```



Better ways of getting some special columns

select()

```{r}
select(flights, year, month, day)

select(arrange(flights, air_time), air_time, origin, dest)
```



That's tedious to write. Hence the pipe.


```{r}
flights %>%
  arrange(air_time) %>%
  select(air_time, origin, dest)
```



Notice that the data doesn't have to be mentioned and the first argument should not have to be provided


```{r}

select(flights, year:day)
# Same as ..

flights %>% select(year:day)
colnames(flights)
```



dropping cols

```{r}
select(flights, -(year:day))
```



Some helper functions

```{r}
select(flights, starts_with("arr"))
select(flights, -starts_with("arr"))
select(flights, ends_with("hour"))
select(flights, -contains("time"))
```

# For more do
?select

# Function for renaming columns
```{r}
rename(flights, destination = dest)
```



Hard to see if it worked, so...
```{r}

flights %>% rename(destination = dest) %>% select(year:day, destination)
```
It is very confusing that the assignment is inversed and counter-intuitive. The column to be renamed is usually on the left hand side of the assignment operator but in this case it is on the right.


Moving some columns to the start

```{r}

select(flights, origin, dest, everything())
```


Class Exercise: What happens if you include a variable multiple times?

ANS: Nothing Happens. The resulting dataframe still displays the column I included multiple times only once.


```{r}

select(flights, origin, origin, origin, dest, dest)
```

Assignment 4

Resources


If you have no experience coding, this may be helpful: https://rstudio-education.github.io/hopr/

Assignment 4

1. Read Chapter 5 of Grolemund and Wickham parts 1 through 4 (including select) of Grolemund and Wickham for anything we did not cover. 
We will cover the remaining parts next week.
2. Turn the script (.R file) from class into a markdown file which displays the graphs and tables. 

Add any comments that might benefit you later on, such as reminders of things you found confusing, etc.
Make sure that you comment the graphs where appropriate, either through captions or in the accompanying text.

3. Repeat the steps from chapter 5 in parts 1 through 3, but using hotels data instead of the nycflights data. 
Since the two datasets don't have the same columns, either pick some variable you'd like to filter on and see results on, or use the following suggested mapping:
Repeat every step for which Grolemund and Wickham show the output - thus ignore all the exercises, or options they mention without.
- When filtering (etc) on month for flights, use stars in the hotels data
- Instead of flight duration, use hotel price
- For travel times, use distance (you can reuse distance for different types of time)

Example: Instead of doing
filter(flights, month == 1)
you should do filter(hotels, stars == <some-number-you-like>)
Create similar output to Grolemund and Wickham, i.e. show what the output is of various commands.

See (Discourse)[https://discourse.trichotomy.xyz/t/week-4-assignment-description/89/2] for an example of what counts as a step.


```{r}

library(tidyverse)
library(scales)

dir <-  Sys.getenv("C:/Users/faaez/OneDrive - Central European University/Current Courses/Data Coding 1 - Data Management and Analysis in R/Assignment 4/R-Coding")

data_in <- paste0(dir,"da_data_repo/hotels-vienna/clean/")
# load vienna
vienna <- read_csv(paste0(data_in,"hotels-vienna.csv"))

```

5.2 Filter

```{r}
filter(vienna, stars == 3)

three_stars <- filter(vienna, stars == 3)

(three_stars <- filter(vienna, stars == 3))
```

5.2.2 Logical Operators

```{r}
filter(vienna, stars == 1 | stars == 3)

#in example
nov_dec <- filter(vienna, month %in% c(11, 12))

#logical operator sample De Morgans Law
#distance less then 2 and stars more than or equal to 3
a <- filter(vienna, !(distance >= 2 | stars < 3))
b <- filter(vienna, distance < 2, stars >= 3)
identical(a, b)
```

5.3 Arrange

```{r}
#arrange example - arrange year, month and then stars in ascending order
arrange(vienna, year, month, stars)


#desc example - arrange by descending stars
arrange(vienna, desc(stars))
```

5.4 Select

```{r}
#select columns by name

select(vienna, city, stars, rating, distance)

#select all columns between country and stars

select(vienna, country:stars)

#select all columns except those between country and stars

select(vienna, -(country:stars))
```

Helper functions examples

```{r}

#starts_with("rating"): 
select(vienna, starts_with("rating"))

#ends_with("city"): 
select(vienna, ends_with("city"))

#contains("rating"): 
select(vienna, contains("rating"))

#matches("(.)\\1"): 
select(vienna, matches("ci"))

#num_range("x", 1:3):
select(vienna, num_range("city", 1:3))
```


Rename Examples

```{r}
rename(vienna, hotel_country = country)
```

select() and everything() helper

```{r}
select(vienna, distance, city, stars, everything())
```

