---
title: "Lecture 2: Chapter 3 by Kieran Healy"
author: "Marc Kaufmann"
date: "January 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Note to the interested student

Try to follow along by typing it yourself, adding comments as you make mistakes or realize things. Write the code out in chunks:

```{r}
```

## How ggplot Works

```{r, load, warning=FALSE}
library(tidyverse)
```

The code specifies the connections between the variables in the data on one hand and the colors, points, and shapes displayed on the screen on the other. These logical connections are called *aesthetic mappings* or simply *aesthetics*.

How to use ggplot:

- `data = gapminder`: Tell it what your data is
- `mapping = aes(...)`: How to map the variables in the data to aesthetics
  - axes, size of points, intensities of colors, which colors, shape of points, lines/points
- Then say what type of plot you want:
  - boxplot, scatterplot, histogram, ...
  - these are called 'geoms' in ggplot's grammar, such as `geom_point()` giving scatter plots
  
  
```{r, geoms, eval=FALSE}
library(ggplot2)
... + geom_point() # Produces scatterplots
... + geom_bar() # Bar plots
... + geom_boxplot() # boxplots
... # 
```

You link these steps by *literally* adding them together with `+` as we'll see.

**Exercise:** What other types of plots are there? Try to find several more `geom_` functions.

## Mappings Link Data to Things You See

```{r}
library(gapminder)
library(ggplot2)
gapminder

p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point()
```

In detail:

- `data = gapminder` tells ggplot to use gapminder dataset, so if variable names are mentioned, they should be looked up in gapminder
- `mapping = aes(...)` shows that the mapping is a function call. Simply accept that this is how you write it
  - Kieran Healy: "The `mapping = aes(...)` argument *links variables* to *things you will see* on the plot"
- `aes(x = gdpPercap, y = lifeExp)` maps the GDP data onto `x`, which is a known aesthetic (the x-coordinate) and life expectancy data onto `x`
  - `x` and `y` are  predefined names that are used by `ggplot` and friends
  
Importantly, mappings don't say *what* color or shape some variable will have. Rather, it says that a given column/variable will be mapped *to* color or *to* shape: thus `color = gender` means that different genders will be displayed by different colors.

```{r}
str(p)
str(p + geom_point())
```

**Exercise:** Make sure that your knitted version doesn't include all the output from the `str(...)` commands, it's too tedious. (Hint: click on the 'gear' symbol in the code block above and then on the link to 'Chunk Options' to get help with chunk options.)

Finally, we add a *layer*. This says how some data gets turned into concrete visual aspects.

```{r, scatter_plot}
p + geom_point()
p + geom_smooth()
```

**Note:** Both geom's use the same mapping, where the x-axis represents ... and the y-axis ... based on the mappings you defined before. But the first one maps the data to individual points, the other one maps it to a smooth line with error ranges.

We get a message that tells us that `geom_smooth()` is using the method = 'gam', so presumably we can use other methods. Let's see if we can figure out which other methods there are.

```{r, smoothing_methods, eval=FALSE}
?geom_smooth
p + geom_point() + geom_smooth() + geom_smooth(method = ...) + geom_smooth(method = ...)
p + geom_point() + geom_smooth() + geom_smooth(method = ...) + geom_smooth(method = ..., color = "red")
```

You may start to see why ggplots way of breaking up tasks is quite powerful: the geometric objects (long for geoms) can all reuse the *same* mapping of data to aesthetics, yet the results are quite different. And if we want later geoms to use different mappings, then we can override them -- but it isn't necessary.

One thing about the data is that most of it is bunched to the left. If we instead used a logarithmic scale, we should be able to spread the data out better.

```{r, scale_coordinates}
p + geom_point() + geom_smooth(method = "lm") + scale_x_log10()
```

**Exercise:** Describe what the `scale_x_log10()` does. Why is it a more evenly distributed cloud of points now? (2-3 sentences.)

Nice! The x-axis now has scientific notation, let's change that.

```{r, scales, eval=FALSE}
library(scales)
p + geom_point() + 
  geom_smooth(method = "lm") + 
  scale_x_log10(labels = scales::dollar)
p + geom_point() + 
  geom_smooth(method = "lm") + 
  scale_x_log10(labels = scales::...)
```

**Exercise:** What does the `dollar()` call do?

```{r, dollar_answer}
?dollar()
```

**Exercise:** How can you find other ways of relabeling the scales when using `scale_x_log10()`?

### The ggplot Recipe

1. Tell the `ggplot()` function what our data is.
2. Tell `ggplot()` *what* variables we want to map to features of plots. For convenience we will put the results of the first two steps in an object called `p`.
3. Tell `ggplot` *how* to display relationships in our data.
4. Layer on geoms as needed, by adding them on the `p` object one at a time.
5. Use some additional functions to adjust scales, labels, tickmarks, titles. 
  - The `scale_`, `labs()`, and `guides()` functions

### Mapping Aesthetics vs Setting them

```{r, mapping_vs_setting}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = 'yellow'))
p + geom_point() + scale_x_log10()
```

This is interesting (or annoying): the points are not yellow. How can we tell ggplot to draw yellow points?

```{r, yellow_points, eval=FALSE}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, ...))
p + geom_point(...) + scale_x_log10()
```

**Exercise:** Based on the discussion in Chapter 3 of *Data Visualization* (read it), describe in your words what is going on. 
One way to avoid such mistakes is to read arguments inside `aes(<property> = <variable>)`as *the property <property> in the graph is determined by the data in <variable>*. 

**Exercise:** Write the above sentence for the original call `aes(x = gdpPercap, y = lifeExp, color = 'yellow')`.

Aesthetics convey information about a variable in the dataset, whereas setting the color of all points to yellow conveys no information about the dataset - it changes the appearance of the plot in a way that is independent of the underlying data. 

Remember: `color = 'yellow'` and `aes(color = 'yellow')` are very different, and the second makes usually no sense, as `'yellow'` is treated as *data* that has the same value always -- namely the value 'yellow'.

```{r, exercise_args_for_smooth}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point() + geom_smooth(color = "orange", se = FALSE, size = 8, method = "lm") + scale_x_log10()
```

**Exercise:** Write down what all those arguments in `geom_smooth(...)` do.

```{r, gapminder_with_labels}
p + geom_point(alpha = 0.3) + 
  geom_smooth(method = "gam") + 
  scale_x_log10(labels = scales::dollar) +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
       title = "Economic Growth and Life Expectancy",
       subtitle = "Data Points are country-years",
       caption = "Source: Gapminder")
```

Coloring by continent:

```{r, gapminder_color_by_continent}
library(scales)
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = continent, fill = continent))
p + geom_point()
p + geom_point() + scale_x_log10(labels = dollar)
p + geom_point() + scale_x_log10(labels = dollar) + geom_smooth(aes(color = continent))

```

**Exercise:** How can you plot to `geom`s for two different datasets on the same plot (assuming they share the same x- and y-axis)? 

```{r, plotting_two_geoms}
## Now plotting two geoms for different data sets

# Get data for points before 1970
gmPre70 <- gapminder %>% filter(year < 1970)
p + geom_point() +
  scale_x_log10(labels = dollar) +
  geom_smooth() +
  geom_smooth(data = gmPre70, color = 'red')

p + geom_point() + scale_x_log10(labels = dollar) + geom_smooth()
```

**Exercise:** What does `fill = continent` do? What do you think about the match of colors between lines and error bands?

```{r, gapminder_color_by_continent_single_line}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + geom_smooth() + scale_x_log10()
```

**Exercise:** Notice how the above code leads to a single smooth line, not one per continent. Why?

**Exercise:** What is bad about the following example, assuming the graph is the one we want? This is why you should set aesthetics at the top level rather than at the individual geometry level if that's your intent.

```{r, many_continents}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp))
p + geom_point(mapping = aes(color = continent)) + 
  geom_smooth(mapping = aes(color = continent, fill = continent)) + 
  scale_x_log10() + 
  geom_smooth(mapping = aes(color = continent), method = "gam")
```

## Additional Exercises 

**Exercise:** Find ways to save the figures that you made so that you can use them elsewhere too. Create a new folder to save only images. Use the command for saving to save the picture for the last image in your new folder, after you have updated the axes, title, subtitle, and caption of the image. Post your solution on Slack and use it to include the final image above with a caption saying "Saved by <your-name>" inside your Slack message (see https://slack.com/help/articles/4403914924435-Add-descriptions-to-images). (Hint: `??save`)

**Exercise:** Read section 3.8 "Where to go next" from DV. Based on those ideas, experiment and create two different graphs with the gapminder data. Describe each briefly in one sentence.

**Exercise:** Read section 1.6 of [R for Data Science](https://r4ds.had.co.nz/introduction.html) on *Getting help and learning more*. Report 3 errors you hit during this week and include a reproducible example of one of these errors as described in that section.

**Exercise:** Do exercise 3.2.4 from [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html#first-steps). Include your code in chunks, describe the output and code (where necessary) in the surrounding text.

**Exercise:** Read the (very short) [Chapter 4 of R for Data Science](https://r4ds.had.co.nz/workflow-basics.html) and try exercise 1 in section 4.4.

**Bonus Exercise:** Why did I load the `scales` library twice via `library(scales)` to knit? 

## Assignment 1 (optional)

1. Do the exercises in these lecture notes.
1. Knit lectures 2 to a pdf file. Get rid of those `eval=FALSE` that are in the current file. They are there to keep RStudio from complaining about code that is incomplete. 